---
title: "Weight Lifting Recognition"
author: "Refik Türkeli"
date: "6/11/2017"
output: html_document
---

```{r include=FALSE}
library(purrr)
library(dplyr)
library(tidyr)
library(knitr)
library(scales)
library(caret)
library(ggplot2)
library(randomForest)
library(rpart)
set.seed(3312)
```

## Introduction

The purpose of this analysis is to recognize common errors that a person does
when performing unilateral dumbbell biceps curl. We have a training dataset
where subjects were asked to perform the excercise while wearing sensor in
5 different fashions which are recorded in the dataset as `classe` variable.
These fashions denote different kinds of errors that are made while performing
the exercise. We will be training a model with the sensor data as input and 
the model will predict the `classe`.

## The Dataset

First lets make some exploratory analysis on the dataset.

```{r cache=TRUE}
dataset <- read.csv("pml-training.csv")
```

The dataset has `r length(colnames(dataset))` variables and `r nrow(dataset)`
observations. Upon inspection, sensors monitor 3 different euler angles:
`roll`, `pitch` and `yaw`. 

There is also a variable called `user_name` which indicates the person that the
sensors record data from. There are `r unique(dataset$user_name)` unique users
and `r unique(dataset$classe)` `classe`.

All data points can be summarised in this `user_name` x `classe` table. Count
of observations are converted to the percentage of entire dataset.

```{r}
dataset %>% 
  group_by(user_name, classe) %>% 
  summarise(count = n()) %>% 
  ungroup() %>% 
  mutate(count = percent(count / sum(count))) %>% 
  spread(classe, count) %>%
  kable()
```

Some variables seem to have a lot of `NA` values. These variables are computed
based on the sensor data. These variables have statistical abbreviations on 
their name like `max`, `stddev` and `avg`. The data is a time series data 
consisting of windows and these computed values exist for the end of windows. 

We will use the variables that start with `avg` to explore the relationship
of sensor data with the `classe`. There are a lot of variables and `avg` would
be a good place to start. Since these data are one per window, we will filter 
out the missing values. Also, I'm normalizing the data points per variable 
to see the characteristic independent of scale.

```{r fig.height=7}
avg_dataset <- dataset %>% 
  filter(new_window == "yes") %>% 
  select(classe, contains("avg_")) %>% 
  gather("colname", "colname_val", contains("avg_")) %>% 
  separate("colname", into = c("method", "angle", "sensor")) %>% 
  select(-method)

ggplot(avg_dataset, aes(classe, colname_val)) +
  geom_point(alpha = 0.1, size = 3) +
  facet_grid(angle ~ sensor, scales = "free_y") +
  ylab("measurement")
```

Some correlation between sensor data and different `classe` is visible. For
example in the case A, arm sensor data seems to be more distributors than in
other cases. For case E, belt yaw values looks closer to each other for above
0 values. For case C, dumbbell roll values seem to have a cluster around -100.
Each of these cases show a different type of error. For each case, measurements
seems to identify a certain characteristics in the movement.

## Variable Selection

For the model selection, we are selecting only features that are not part
of a statistics like `mean` or `max`. Those values are once for repetition and
mostly `NA`. Our model should be able to work in their absence.
Also I'm excluding the `user_name` variable, because the model has to be 
generic and work for anyone. 

```{r}
cols <- names(dataset)
statistic_abbrev <- c("avg", "stddev", "skewness", "var", "max", "min",
                      "kurtosis", "total", "amplitude")
nonstat_cols <- cols[reduce(map(statistic_abbrev, ~ !grepl(.x, cols)), `&`)]

dataset_nostat <- dataset %>%
  filter(new_window != "yes") %>%
  select_(.dots = nonstat_cols) %>%
  select(-contains("timestamp"), -new_window, -X, -user_name)
```

Making a PCA to determine variables that make up the 80% of the variation
in the dataset.

```{r}
pca_pp <- preProcess(dataset_nostat %>% select(-classe), 
                     thresh = 0.8, method = "pca")

pca_pp
```

## Partitioning Data

We should partition the before training our model to get a better estimate
of our out of sample error rate.

```{r}
in_train <- createDataPartition(dataset_nostat$classe, p = 0.7, list = FALSE)
training <- dataset_nostat[in_train,]
testing <- dataset_nostat[-in_train,]
```

## Fitting Models

I will start by fitting a decision tree. This model will use k-fold cross
validation to improve the fit.

```{r}
ctree <- train(classe ~ ., data = training, method = "rpart",
               trControl = trainControl(method = "cv"))
ctree
```

Accuracy of the classification tree is not great. I will try random forests
with bootstrap aggregating error measurement in order to select the best fit.

```{r cache=TRUE}
rf <- train(classe ~ ., data = training, method = "rf",
            trControl = trainControl(method = "oob"))
rf
```

Random forests seem to have good accuracy based on cross validation. Now I will
apply it once on the test set to get a better estimate for the out of sample
error.

```{r}
predictions <- predict(rf, newdata = testing %>% select(-classe))
confusionMatrix(predictions, testing$classe)
```

## Conclusion

Random forest model with the test set prediction indicates similar accuracy 
to cross validation with 99%. This seems to be a very successful model.